Proportional multi-state multiple-cohort life table model

### 2018-04-23 MSLT/MSLTdoc.Rmd
Generated general life tables

7) Create age and sex categories to obtain population numbers. Population numbers from GBD are used in Dismod II. For the Life table (Figure 3), the numbers may be from the synthetic population. For now, the Life table is set up with population numbers derived from the GBD data. 
```{r, echo=TRUE}
GBDdata$sex_age_cat <- paste(GBDdata$sex,GBDdata$age_cat, sep = "_"  )
```
8) Conver string variables to numberic to do calculations.
```{r, echo=TRUE}
GBDdata$val <- as.numeric(as.character(GBDdata$val))
```
9) Generate population numbers for Greater London in a new data frame ("GBD_population"). Note that there is data for England as well, which is used in a separate rmarkdown dowcument (GBDCompare). Population numbers are derived from rates per 100,000 and total numbers of cases. 
```{r, echo=TRUE}
GBD_population <- filter(GBDdata, measure == "deaths", cause == "all causes", metric == "rate" | metric == "number" ) %>% select(metric, age_cat, sex, val, sex_age_cat, location)
```
10) Generate population numbers from given number of cases and rates per 100,000 people. 
```{r, echo=TRUE}
 for (i in 1:nrow(GBD_population)) {
    if (GBD_population$metric[i] == "number") {
      GBD_population$pop_five_year[i] <- GBD_population$val[i] * 100000/ GBD_population$val[i + 2]}
   else {GBD_population$pop_five_year[i] <- NA}
 }
```
11) Remove rows with zero
```{r, echo=TRUE}
GBD_population <- GBD_population[!is.na(GBD_population$pop_five_year),]
```
12) Keep relevant variables
```{r, echo=TRUE}
GBD_population <- filter(GBD_population) %>% select(sex_age_cat, sex, age_cat, pop_five_year, location)
```
13) Create data frames for Greater London to be later used for: a) interpolation of rates and b) PMLT cohorts. 
```{r, echo=TRUE}
GBD_population_GL <-  filter(GBD_population, location == "greater london") %>% select(sex_age_cat, age_cat, sex, pop_five_year, location)
```
12) Check population total numbers
```{r, echo=TRUE}
GreaterLondon <- sum(GBD_population_GL$pop_five_year)
```

@@

interpolated <- as.data.frame(interpolation_func(seq(0, 99, 1)))
ages <- seq(0,99,by=1)
interpolated <- cbind(interpolated, ages)
interpolated[,1] <- exp(interpolated[,1])
colnames(interpolated)[1] <- "rate"

assign(paste(sex_index, measure_index, "interpolated", sep = "_"), interpolated)
# plot(interpolated$age,interpolated$rate)

##Do graph with another layer for original rates in age groups for comparison purposes.
p <- ggplot(data = interpolated,mapping = aes(ages,rate)) +
              geom_line(aes(color = "Interpolated")) +
              geom_point(
              data = data, 
              mapping = aes(age_cat, one_rate, color = "Original")) +
              labs(colour="",x="Age",y= paste(i_sex, i_measure, "rates", sep = " "))+
            labs (title = paste("Rates", i_sex, i_measure, sep = " "), size=14) +
            theme_classic() +
            theme(plot.title = element_text(hjust = 0.5))
            print(p)
  }
}

@@

```{r, echo=TRUE}
##Create life table input data frame (population numbers, all-cause moratlity rates ##(per one) and all cause yld rates)

##Create empty data frame
input_life_table <- as.data.frame(matrix(0, ncol=4, nrow = 202))
##Give variables names (mx=total mortality rate, yldsx = total years lived with disability rates )
names(input_life_table) <- c("age", "sex", "mx", "yldsx")
##Populate life table: Age, 0 to 100
input_life_table[1:101, 1] <- c(0:100)
input_life_table[102:202, 1] <- c(0:100)
##Populate life table: sex, male, female
input_life_table[1:101, 2] <- "male"
input_life_table[102:202, 2] <- "female"
##Create variable age_sex to match with population data. 
input_life_table$sex_age_cat <- paste(input_life_table$sex,input_life_table$age, sep = "_"  )

##Populate life table: five_year_pop. SEE LIFE TABLE LOOPS THAT USES 5 YEAR POP.

input_life_table <- merge(input_life_table, select(GBD_population_GL
                                                     ,c(sex_age_cat, pop_five_year))
                                                     , by = 'sex_age_cat', all.x = TRUE)

##Cross check population numbers
sum(input_life_table$pop_five_year, na.rm = TRUE)
#Sort data by sex and age to use in life table function. 
input_life_table<-input_life_table[order(input_life_table$sex, input_life_table$age),]

```
Health outcomes included in leisure time meta-analysis: all-cause mortality, cardiovascular disease, stroke, coronary heart disease, total cancer, colon, lung, endometrial, and breast cancer.

>>>>

7) Create age and sex categories to obtain population numbers. Population numbers from GBD are used in Dismod II. For the Life table (Figure 3), the numbers may be from the synthetic population. For now, the Life table is set up with population numbers derived from the GBD data. 
```{r, echo=TRUE}
GBDdata$sex_age_cat <- paste(GBDdata$sex,GBDdata$age_cat, sep = "_"  )
```
8) Conver string variables to numberic to do calculations.
```{r, echo=TRUE}
GBDdata$val <- as.numeric(as.character(GBDdata$val))
```
9) Generate population numbers for Greater London in a new data frame ("GBD_population"). Note that there is data for England as well, which is used in a separate rmarkdown dowcument (GBDCompare). Population numbers are derived from rates per 100,000 and total numbers of cases. 
```{r, echo=TRUE}
GBD_population <- filter(GBDdata, measure == "deaths", cause == "all causes", metric == "rate" | metric == "number" ) %>% select(metric, age_cat, sex, val, sex_age_cat, location)
```
10) Generate population numbers from given number of cases and rates per 100,000 people. 
```{r, echo=TRUE}
 for (i in 1:nrow(GBD_population)) {
    if (GBD_population$metric[i] == "number") {
      GBD_population$five_year_population[i] <- GBD_population$val[i] * 100000/ GBD_population$val[i + 2]}
   else {GBD_population$five_year_population[i] <- NA}
 }
```
11) Remove rows with zero
```{r, echo=TRUE}
GBD_population <- GBD_population[!is.na(GBD_population$five_year_population),]
```
12) Keep relevant variables
```{r, echo=TRUE}
GBD_population <- filter(GBD_population) %>% select(sex_age_cat, sex, age_cat, five_year_population, location)
```
13) Create data frames for Greater London to be later used for: a) interpolation of rates and b) PMLT cohorts. 
```{r, echo=TRUE}
GBD_population_GL <-  filter(GBD_population, location == "greater london") %>% select(sex_age_cat, age_cat, sex, five_year_population, location)
```
12) Check population total numbers
```{r, echo=TRUE}
GreaterLondon <- sum(GBD_population_GL$five_year_population)
```

@@

interpolated <- as.data.frame(interpolation_func(seq(0, 100, 1)))
age <- seq(0,100,by=1)
interpolated <- cbind(interpolated, age)
interpolated[,1] <- exp(interpolated[,1])
colnames(interpolated)[1] <- paste(measure_index)
# Add column with sex to create age_sex category to then merge with input_life table
interpolated$sex <- paste(sex_index)
interpolated$sex_age_cat <- paste(interpolated$sex, interpolated$age, sep = "_")
# Change name of column death to mx and ylds to pyld_rate to then merge with input_life table

if (colnames(interpolated)[1] == "deaths")
    colnames(interpolated)[1] <- paste("mx")
  else
    colnames(interpolated)[1] <- paste("pyld_rate")

# Name data frame
assign(paste(sex_index, measure_index, "interpolated", sep = "_"), interpolated)
# plot(interpolated$age,interpolated$rate)

##Do graph with another layer for original rates in age groups for comparison purposes.
p <- ggplot(data = interpolated,mapping = aes(age, interpolated[,1])) +
              geom_line(aes(color = "Interpolated")) +
              geom_point(
              data = data, 
              mapping = aes(age_cat, one_rate, color = "Original")) +
              labs(colour="",x="Age",y= paste(i_sex, i_measure, "rates", sep = " "))+
            labs (title = paste("Rates", i_sex, i_measure, sep = " "), size=14) +
            theme_classic() +
            theme (plot.title = element_text(hjust = 0.5))
            print(p)
  }
}

@@

```{r, echo=TRUE}
##Create life table input data frame (population numbers, all-cause moratlity rates ##(per one), all cause yld rates) and 5-year age groups 
## to run the cohorts (from population life table).

##Create empty data frame
input_life_table <- as.data.frame(matrix(0, ncol=2, nrow = 202))
##Give variables names (mx=total mortality rate, yldsx = total years lived with disability rates )
names(input_life_table) <- c("age", "sex")
##Populate life table: Age, 0 to 100
input_life_table[1:101, 1] <- c(0:100)
input_life_table[102:202, 1] <- c(0:100)
##Populate life table: sex, male, female
input_life_table[1:101, 2] <- "male"
input_life_table[102:202, 2] <- "female"
##Create variable age_sex to match with population data. 
input_life_table$sex_age_cat <- paste(input_life_table$sex,input_life_table$age, sep = "_"  )

##Populate life table: five_year_pop.

input_life_table <- merge(input_life_table, select(GBD_population_GL
                                                     ,c(sex_age_cat, five_year_population))
                                                     , by = 'sex_age_cat', all = TRUE) 
##Populate life table: mortality rates (from interpolated rates)
## Males mortality rates. NEED MATCHING NAMES
input_life_table <- merge(input_life_table, select(male_deaths_interpolated
                                                     ,c(sex_age_cat, mx))
                                                     , by = 'sex_age_cat', all.x = TRUE)
input_life_table <- merge(input_life_table, select(female_deaths_interpolated
                                                     ,c(sex_age_cat, mx))
                                                     , by = 'sex_age_cat', all.x = TRUE)
input_life_table <- merge(input_life_table, select(male_ylds_interpolated
                                                     ,c(sex_age_cat, pyld_rate))
                                                     , by = 'sex_age_cat', all.x = TRUE) 
input_life_table <- merge(input_life_table, select(female_ylds_interpolated
                                                     ,c(sex_age_cat, pyld_rate))
                                                     , by = 'sex_age_cat', all.x = TRUE)

## I created mx.x and mx.y and pyld.x and pyld.y and I only need mx and pyld_rate in one colument. I wil do a quick fix here. 
## DISCUSS WITH ALI TO IMPROVE

input_life_table$mx <-  ifelse(input_life_table$sex == "male", input_life_table$mx.x, input_life_table$mx.y)
input_life_table$pyld_rate <-  ifelse(input_life_table$sex == "male", input_life_table$pyld_rate.x, input_life_table$pyld_rate.y)

##Cross check population numbers
sum(input_life_table$five_year_population, na.rm = TRUE)
#Sort data by sex and age to use in life table function. 
input_life_table<-input_life_table[order(input_life_table$sex, input_life_table$age),]

```
16) Generate baseline life tables from input_data_frame. Life tables are generated for each age (5-years) and sex cohort. 
It is assumed that current observed rates of mortality and disability are going to be observed in the future.

``` {r, echo=TRUE}

##Generate list of life tables. 
## The function run_life_table was created to run the basline and scenario life tables (see functions.R script). 
## The input_life_table data frame was generated to work with the function.

## Generate looping variables (sex and age_group. Sex already generated in code for interpolation (i_sex))

i_age_cohort <- c(22, 27, 32, 37, 42, 47, 52, 57, 62, 67, 72, 77, 82, 87, 92, 97)

general_life_table_list_bl <- list()

index <- 1

for (age in i_age_cohort){
  for (sex in i_sex){
    cat("age ", age, " and sex ", sex, "\n")
    general_life_table_list_bl[[index]] <- run_life_table(in_idata = input_life_table, in_sex = sex, in_mid_age = age)
    index <- index + 1
  }
}

##### Uncommnet to check life table list
View(general_life_table_list_bl[[2]])

17) Generate disease life tables. 
Generate two sets for Greater London: 1) GBD data and 2) English survey data.

Health outcomes included in leisure time meta-analysis: all-cause mortality, cardiovascular disease, stroke, coronary heart disease, total cancer, colon, lung, endometrial, and breast cancer.
